---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "inst/img/"
)
```

# mrmr2: mark recapture miscellany in R, extended

![build status](https://github.com/SNARL1/mrmr2/workflows/R-CMD-check/badge.svg)

This package automates common data processing steps for mark recapture data,
with an emphasis on data collected by researchers in the [`Mountain Lakes Research Group`](https://mountainlakesresearch.com) (University of California, Sierra Nevada Aquatic Research Laboratory).
Many of the assumptions made in the data processing and modeling are pinned to
the type of capture-recapture data collected by the SNARL team, and as a result,
this package should not be thought of as a general purpose mark-recapture
modeling toolbox (for that, see the RMark, unmarked, multimark or Rcapture
packages).

`mrmr2` extends the original [`mrmr`](https://github.com/SNARL1/mrmr) package
with two optional detection-heterogeneity random-effect models and a
lower-memory survival plotting method, and can be installed alongside `mrmr`
without conflict.

## Installation

```{r, eval = FALSE}
remotes::install_github("SNARL1/mrmr2")
```

## Usage

This package consists of three core functions to use consecutively:

1. `clean_data()` ingests capture-recapture, introduction, and survey data.
2. `fit_model()` fits a Jolly-Seber mark recapture model to quickly estimate
demographic parameters and population abundance. By default this is the
same model used by `mrmr`, but the `model` argument can request one of two
optional detection random-effect extensions (`"detect_re"` for
between-period/occasion detection variation, `"indiv_re"` for individual-level
detection heterogeneity) when unmodeled detection variation is a concern.
3. `plot_model()` generates plots of abundance and recruitment over time, or
of the survival of introduced cohorts. Its survival plot pulls posterior
draws only for the individuals it needs, keeping memory use low even for
sites with a large augmented superpopulation.

For an in-depth look at how to use these functions, see the vignette
[An Introduction to mrmr2](https://snarl1.github.io/mrmr2/articles/intro-to-mrmr2.html). Additional details on the new functionality added to mrmr2 is provided in [An Introduction to CMR Analysis Using the mrmr and mrmr2 Packages](https://github.com/SNARL1/mrmr-revisited/blob/main/introduction-to-cmr-with-mrmr.qmd). 

## Background

Customized mark recapture modeling can be time and resource intensive, 
especially when there are unique features of a study system that prevent or 
complicate the immediate use of a general purpose package. 
For example, the model described in Joseph and Knapp 2018 **Ecosphere** 
(https://doi.org/10.1002/ecs2.2499) - which accounts for introductions into a
population, and continuous, time-varying, incompletely observed individual-level
covariates on detection and survival - takes considerable effort to develop and
deploy. 
There is a need in this particular research group for models that can be used 
in near-real time as data are available to inform conservation decisions. 
This package fills this need, with the primary goal of estimating abundance 
and recruitment through time in montane amphibian populations. 

## License

This package is licensed under the MIT license. 

Copyright 2019 Maxwell B. Joseph

Permission is hereby granted, free of charge, to any person obtaining a copy of 
this software and associated documentation files (the "Software"), to deal in 
the Software without restriction, including without limitation the rights to 
use, copy, modify, merge, publish, distribute, sublicense, and/or sell copies 
of the Software, and to permit persons to whom the Software is furnished to do 
so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all 
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE 
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER 
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, 
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE 
SOFTWARE.