---
title: "An introduction to mrmr2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to mrmr2}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5
)
```

The `mrmr2` package is designed to simplify and expedite mark recapture
data processing, parameter estimation, and visualization for a narrow
class of mark recapture data (namely, those collected by researchers in
the Mountain Lakes Research Group (University of California, Sierra
Nevada Amphibian Research Laboratory). It extends the original `mrmr`
package with optional detection-heterogeneity random effects (see
[Detection random-effect models](#detection-re)) and a lower-memory
survival plotting method, and can be installed alongside `mrmr` without
conflict.

# Workflow overview

The typical workflow involves three steps:

1.  Cleaning mark recapture data with `mrmr2::clean_data()`
2.  Fitting a mark recapture model with `mrmr2::fit_model()`
3.  Visualizing results with `mrmr2::plot_model()`

## Data specifications

The package comes with example data that illustrate the expected format.
Three files are needed. These files are typically named something like
`capture`, `translocation`, and `survey`, but the files are specified in
the [data processing step](#data-process) and any names are acceptable.

The first file specifies capture data:

```{r setup}
library(dplyr)
library(readr)
library(tibble)
library(mrmr2)
```

```{r show-capture-data}
capture <- read_csv(
  system.file(
    "extdata", 
    "capture-example.csv",
    package = "mrmr2"
  )
)
capture <- capture |> 
mutate(pit_tag_id = as.character(pit_tag_id))

glimpse(capture)
```

Note that the `system.file` function generates a path to the
`capture-example.csv` file, but in a real application, you would use a
path to a local file (real data, not the example data included in the
package). The capture file must contain, at a minimum, `pit_tag_id` and
`survey_date` columns, but additional columns can also be included
(e.g., [survival covariates](#survival-covar)). Each row in this file
represents a capture or recapture event.

The second file specifies translocation/reintroduction data, and at a
minimum it must contain `release_date` and `pit_tag_id` columns:

```{r show-trans-data}
translocation <- read_csv(
  system.file(
    "extdata", 
    "translocation-example.csv", 
    package = "mrmr2"
  )
)

glimpse(translocation)
```

Each row corresponds to a translocation event of one unique individual.
Because some mark-recapture datasets do not include translocation or
reintroduction actions, this file is optional.

The third file specifies survey-level data, and at a minimum it must
contain `survey_date`, `primary_period`, and `secondary_period columns`:

```{r show-survey-data}
survey <- read_csv(
  system.file(
    "extdata", 
    "survey-example.csv", 
    package = "mrmr2"
  )
)

glimpse(survey)
```

Here, each row is a unique survey. If translocations or reintroductions
occurred at the site, these actions need to be included in the `survey`
dataset, each as a typical survey event. Specifically, `survey_date` and
`primary_period` values are provided, the `primary_period` is sequential
with all survey records in the dataset based on `visit_date`, and
`secondary_period` = 0. Translocations/reintroductions cannot occur
during a primary period. Therefore, the `survey_date` for
translocation/reintroduction records cannot overlap with the
`survey_date` range of any primary period.

## Data processing {#data-process}

To load and process the data, provide paths to the data files as
arguments to the `mrmr2::clean_data()` function:

```{r clean-data, message=FALSE}
data <- clean_data(capture, survey, translocation)
```

This function performs several data-preparation steps. It checks the
capture, survey, and translocation/reintroduction data for common
problems (e.g., duplicate records, dead animals left in the capture
data, unparseable dates), and renumbers the primary periods and inserts
a placeholder pre-study period, deriving each period's representative
survey date, year, and overwinter status from the provided dates. It
builds the individual-by-primary-period-by-secondary-period capture
history, then performs data augmentation: padding the observed
individuals with a large number of all-zero pseudo-individuals to form
the fixed-size superpopulation of $M$ individuals referenced in [Model
structure](#model-structure) below. Finally, it constructs the detection
and survival covariate design matrices from the `capture_formula` and
`survival_formula` arguments, described in the next two sections. The
output provides data frames as list elements, and a list element called
`stan_d`, which contains data that has been pre-processed for model
fitting.

### Including detection/non-detection covariates

Survey-level covariates that might affect the probability of detection
can be included via the `capture_formula` argument to `clean_data`.
**Note:** it is a good idea to ensure that any continuous covariates are
rescaled prior to using them as covariates to have mean = 0 and unit
standard deviation. This can be achieved by adding a column to the
`captures` data Using a covariate value like `year`, which typically has
a mean in the thousands and a standard deviation far from one may cause
numerical issues when fitting a model.

A detection covariate, such as `person_hours`, can be included as
follows:

```{r detection-covariate, message=FALSE, eval=FALSE}
# center and scale the covariate
captures <- captures |>
  mutate(c_person_hours = c(scale(person_hours)))

# specify the formula
data <- clean_data(captures, surveys, translocations,
                   capture_formula = ~ c_person_hours)
```

### Including survival covariates {#survival-covar}

Individual-level survival covariates can be included via the
`survival_formula` and `survival_fill_value` arguments. Both of these
arguments must be specified because covariate values for
pseudo-individuals must be filled in (they are never observed). For
example, to evaluate the effect of an experimental treatment, if some
individuals belong to a "treated" group and others belong to an
"control" group, then one of these groups must be specified as a fill
value (e.g., "treated" or "control", depending on the experiment):

```{r survival-covariate, message=FALSE, eval=FALSE}
# specify the formula
data <- clean_data(captures, surveys, translocations,
                   survival_formula = ~ treatment, 
                   survival_fill_value = c(treatment = "treated"))
```

## Model structure {#model-structure}

The `mrmr2` package implements a Bayesian open-population Jolly-Seber
mark recapture model with known additions to the population (introduced
adults). The model tracks the states of $M$ individuals that comprise a
superpopulation made up of real and pseudo-individuals (see Joseph and
Knapp 2018, *Ecosphere* for details).

We assume a robust sampling design where the states of individuals are
constant within primary periods, which have repeat secondary periods
within them (during which observations are made). The possible states of
individuals include "not recruited", "alive", and "dead". The possible
observations of individuals include "detected" and "not detected". We
assume that individuals that are in the "not recruited" or "dead" states
are never detected (i.e., there are no mistakes in the individual PIT
tag ID records).

## Model fitting

To fit a mark-recapture model, use `mrmr2::fit_model()`. This model
accounts for known introductions into the population, and has random
effects to account for variation in survival and recruitment through
time. It is also possible to add random effects to account for
[variation in detection](#detection-re) at the period, occasion, and
individual level.

The default number of chains is 4. Generally at least 1000 warmup and
1000 sampling iterations should be used; the example model below uses
only 10 iterations of each to ensure a short run time.

```{r fit-model, message=FALSE, warning=FALSE, results='hide'}
mod <- fit_model(
  data = data, 
  chains = 4,
  iter_warmup = 10, 
  iter_sampling = 10
)
```

To use within-chain parallelization, provide the following additional
arguments:

- `cpp_options = list(stan_threads = TRUE)` which compiles the model
  with threading support
- `threads_per_chain` set to some integer value \> 1. To allow the run
  to finish as quickly as possible, set this to a value that ensures
  that all physical cores are being used. The number of parallel_chains
  × threads_per_chain must equal the number of physical cores available.
  For example, on a machine with 8 physical cores, specify 4 chains, 4
  parallel chains, and 2 threads per chain. On a machine with 16
  physical cores, specify 4 chains, 4 parallel chains, and 4 threads per
  chain. In the following code chunk, threads per chain is set
  automatically.

```{r fit-model-threaded, message=FALSE, warning=FALSE, eval=FALSE}
mod <- fit_model(
  data = data, 
  chains = 4,
  parallel_chains = 4,
  threads_per_chain = max(1, parallel::detectCores() %/% 4), 
  cpp_options = list(stan_threads = TRUE),
  iter_warmup = 1000, 
  iter_sampling = 1000
)
```

To diagnose potential non-convergence of the MCMC algorithm, inspect
traceplots:

```{r plot-traceplots}
pars_to_plot <- c('alpha_lambda', 
                  'sigma_lambda', 
                  'beta_phi', 
                  'sigma_phi', 
                  'beta_detect')
bayesplot::mcmc_trace(mod$m_fit$draws(pars_to_plot))
```

Also check the Rhat estimates to see whether they indicate of a lack of
convergence (Rhat values $\geq$ 1.01):

```{r print-model}
mod$m_fit$summary(pars_to_plot)
```

### Detection random-effect models {#detection-re}

By default (`model = "base"`), detection probability is a purely
deterministic function of any covariates supplied via `capture_formula`
to `clean_data()` — with the default `capture_formula = ~ 1`, every
survey in the dataset shares one scalar detection probability. Survival
and recruitment, by contrast, already have a random effect that lets
them vary by primary period. `mrmr2::fit_model()` offers two optional
extensions, selected with the `model` argument, that add an analogous
random effect to detection:

- `model = "detect_re"` adds a between-primary-period random effect and
  a within-primary-period (per-survey) random effect to detection
  probability. Simulation work found this mainly corrects overconfident
  (too-narrow) uncertainty intervals on abundance (`N`) and recruitment
  (`B`) when true detectability drifts across periods or survey
  occasions — it has little effect on the point estimates themselves.
- `model = "indiv_re"` adds an individual-level random effect, so that
  some individuals can be persistently harder to detect than others
  (e.g., due to body size, behavior, or microhabitat use). Unlike
  `"detect_re"`, ignoring real individual heterogeneity of this kind
  causes a genuine *negative bias* in point estimates of `N` —
  chronically hard-to-detect individuals get undercounted — matching
  classical closed-population heterogeneity (`M_h`) theory.

```{r fit-model-re, eval=FALSE}
mod_indiv_re <- fit_model(
  data = data,
  model = "indiv_re",
  parallel_chains = parallel::detectCores(),
  iter_warmup = 1000,
  iter_sampling = 1000
)
```

These extensions are not free. Both take longer to fit than `"base"`,
and `"indiv_re"` in particular produces substantially larger saved fits,
since its random effect has one entry per individual in the augmented
superpopulation (`M`, typically several times the number of animals ever
observed) rather than one per primary period or survey. `"detect_re"`'s
two variance components are only weakly separable, which can trigger
divergent transitions when one of them is at or near zero; `fit_model()`
handles this automatically by defaulting `adapt_delta` to `0.95` (rather
than cmdstanr's usual `0.8`) whenever `model = "detect_re"` and
`adapt_delta` is left unspecified.

One more caveat when interpreting a `"detect_re"` fit: with only a few
secondary surveys per primary period, the model can attribute some of
what is really per-survey (`sigma_detect_occasion`) noise to the
between-period term (`sigma_detect_period`) instead, purely because a
small sample of occasion-level draws can look like a period-level shift
by chance. Before concluding there is genuine period-to-period drift in
detectability, compare the posterior for `sigma_detect_period` against
the noise floor it would show even with no real period effect, roughly
`sigma_detect_occasion / sqrt(J)`, where `J` is the number of secondary
surveys in a primary period. This labeling ambiguity doesn't bias `N`,
`B`, `phi`, or `lambda` — those depend only on the sum of the two random
effects — but it does mean the two `sigma` estimates shouldn't be read
in isolation.

## Built-in visualizations

Time series of abundance, recruitment, and survival of introduced
cohorts are available through the `mrmr2::plot_model()` function.

```{r plot-abundance}
plot_model(mod, what = 'abundance')
```

```{r plot-recruitment}
plot_model(mod, what = 'recruitment')
```

```{r plot-survival}
plot_model(mod, what = 'survival')
```

`what = 'survival'` only pulls posterior draws of the latent state (`s`)
for translocated individuals, rather than for every individual (real and
data-augmented pseudo-individual) tracked by the model. On sites with a
large superpopulation relative to the number of translocated
individuals, this keeps memory use for this plot far below what it would
be if the full `s` array were pulled and then filtered down afterward.

### Translocation survival tables

If individuals have been translocated into sites, you can generate a
table of estimated survival through time for each cohort (release date)
as a function of years since introduction:

```{r survival-table}
survival_table(mod)
```

You can also get estimates of survival through time by individual or
cohort:

```{r survival-table-cohort}
survival_table(mod, by_cohort = TRUE)
```

```{r survival-table-ind}
survival_table(mod, by_individual = TRUE)
```

## Custom visualizations

Any of the plotting functionality that you would expect from a `stanfit`
model is available as well, by accessing the `m_fit` list element from a
model object. For example, we could assess the posterior for the
superpopulation size:

```{r}
bayesplot::mcmc_trace(mod$m_fit$draws("Nsuper"))
```

## Cases with no translocations

In some cases, there may not be any introductions of animals into a
population. The only difference in implementation in such cases is that
the `translocations` argument to `clean_data()` can be omitted:

```{r no-trans, message = FALSE, eval = FALSE}
data_no_trans <- clean_data(captures, surveys)
mod_no_trans <- fit_model(data_no_trans)
```

## Cases with known removals

When individuals are removed from a population, e.g., for translocation,
these known removals can be included as data to increase the precision
of abundance estimates. An example dataset fitting this case is included
at an imaginary site called "Equid". Note below that we filter out dead
animals -- any dead animals encountered on surveys should be in the
`removals` data, and not in the `captures` data frame, which should only
contain live animals (those that might survive to the next primary
period).

```{r removals, message = FALSE, eval = FALSE}
captures <- system.file("extdata", 
                        "equid-captures.csv", 
                         package = "mrmr2") |>
  read_csv |>
  filter(capture_animal_state != 'dead')
surveys <- system.file("extdata", 
                       "equid-surveys.csv", 
                       package = "mrmr2") |>
  read_csv
removals <- system.file("extdata", 
                        "equid-removals.csv", 
                         package = "mrmr2") |>
  read_csv

d <- clean_data(captures, surveys, removals = removals)
m <- fit_model(d, parallel_chains = 2, chains = 2)
```
