---
title: "An introduction to mrmr2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to mrmr2}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5
)
```

The `mrmr2` package is designed to simplify and expedite mark-recapture
data processing, parameter estimation, and visualization for a narrow
class of mark-recapture data (namely, those collected by researchers at
the [Mountain Lakes Research Group](https://mountainlakesresearch.com)
(University of California, [Sierra Nevada Aquatic Research
Laboratory](https://snarl.nrs.ucsb.edu/)). It extends the original
`mrmr` package with optional detection-heterogeneity random effects (see
[Detection random-effect models](#detection-re)) and a lower-memory
survival plotting method, and can be installed alongside `mrmr` without
conflict. `mrmr2` also incorporates numerous code fixes and improved
default plots.

## Install latest versions of cmdstanr, cmdstan, and mrmr2 packages

The following code chunk only installs/updates a package if it's missing
or outdated -- an already up-to-date package is left alone.

```{r install-stan-mrmr2-packages, eval=FALSE}
# install/update cmdstanr from its dedicated repository
remotes::update_packages("cmdstanr",
                         repos = c("https://mc-stan.org/r-packages/", getOption("repos")))

# install/update CmdStan itself
cmdstanr::install_cmdstan()

# install/update mrmr2 from GitHub
remotes::install_github("SNARL1/mrmr2")
```

# Workflow overview

The typical workflow involves three steps:

1.  Cleaning mark-recapture data with `mrmr2::clean_data()`
2.  Fitting a mark-recapture model with `mrmr2::fit_model()`
3.  Visualizing results with `mrmr2::plot_model()`

## Data specifications

The package comes with example data that illustrate the expected format.
Two files are required (describing captures and surveys) and two files
are optional (describing additions and removals). These files are
specified in the [Data processing](#data-process) step and are typically
named something like `capture`,`survey`, `translocation`, and `removal`,
but any names are acceptable.

The first file specifies capture data:

```{r setup}
library(dplyr)
library(readr)
library(tibble)
library(mrmr2)
library(ggplot2)
library(patchwork)
```

```{r show-capture-data}
capture <- read_csv(
  system.file(
    "extdata", 
    "capture-example.csv",
    package = "mrmr2"
  )
)
capture <- capture |> 
mutate(pit_tag_id = as.character(pit_tag_id))

glimpse(capture)
```

Note that the `system.file` function generates a path to the
`capture-example.csv` file, but in a real application, you would use a
path to a local file (real data, not the example data included in the
package). The capture file must contain, at a minimum, `pit_tag_id` and
`survey_date` columns, but additional columns can also be included
(e.g., [survival covariates](#survival-covar)). Each row in this file
represents a capture event, and only live individuals are included. Any
previously-tagged individuals found dead during during surveys can be
included in the `removal` file (see below).

The second file specifies survey-level data, and at a minimum it must
contain `survey_date`, `primary_period`, and `secondary_period` columns:

```{r show-survey-data}
survey <- read_csv(
  system.file(
    "extdata", 
    "survey-example.csv", 
    package = "mrmr2"
  )
)

glimpse(survey)
```

Each row is a unique survey. If translocations or reintroductions
occurred at the site, these actions need to be included in the `survey`
dataset, each as a typical survey event. Specifically, `survey_date` and
`primary_period` values are provided, the `primary_period` value is
sequential with all survey records in the dataset based on `visit_date`,
and `secondary_period` = 0. Translocations/reintroductions cannot occur
during a primary period. Therefore, the `survey_date` for
translocation/reintroduction records cannot overlap with the
`survey_date` range of any primary period.

The third file specifies translocation/reintroduction data, and at a
minimum it must contain `release_date` and `pit_tag_id` columns:

```{r show-trans-data}
translocation <- read_csv(
  system.file(
    "extdata", 
    "translocation-example.csv", 
    package = "mrmr2"
  )
)

glimpse(translocation)
```

Each row is an addition event of one unique individual. Because some
mark-recapture datasets do not include translocations or
reintroductions, this file is optional.

The fourth file specifies removal data and, at a minimum, contains,
`removal_date` and `pit_tag_id` columns. Each row is a removal event of
one unique individual, including previously-tagged individuals collected
for translocation to other sites and previously-tagged individuals found
dead during surveys. Because some mark-recapture datasets do not include
removals, this file is optional.

```{r show-remove-data}
removal <- read_csv(
  system.file(
    "extdata", 
    "removal-example.csv", 
    package = "mrmr2"
  )
)

glimpse(removal)
```

## Data processing {#data-process}

To load and process the data, provide the data frames as named arguments
to the `mrmr2::clean_data()` function: `captures`, `surveys`,
`additions` (translocations or reintroductions, optional), and
`removals` (optional). Note that these are the names `clean_data()`
requires for its arguments -- they need not match the names you gave
your own data frames. For example, in the example below, the data frame
`translocation` (a name we chose when reading in the file above) is
passed as `clean_data()`'s `additions` argument. `clean_data()`
determines the role of each input solely from these argument names, and
will raise an error if any argument is passed positionally instead of by
name (e.g., `capture` instead of `captures = capture`):

```{r clean-data, message=FALSE}
data <- clean_data(
  captures = capture, 
  surveys = survey,
  additions = translocation, 
  removals = removal
  )
```

In some cases, there may be no introductions or removals of individuals
into/from a population. In such cases, the only difference in
implementation is that the `additions` and/or `removals` arguments to
`clean_data()` are omitted.

The `mrmr2:clean_data` function performs several data-preparation steps.
It checks the capture, survey, translocation/reintroduction, and removal
data for common problems (e.g., duplicate records, dead animals left in
the capture data, unparseable dates), and renumbers the primary periods
and inserts a placeholder pre-study period, and derives each period's
representative survey date, year, and overwinter status from the
provided dates. It builds the
individual-by-primary-period-by-secondary-period capture history, then
performs data augmentation: padding the observed individuals with a
large number of all-zero pseudo-individuals to form the fixed-size
superpopulation of $M$ individuals referenced in [Model
structure](#model-structure) below. Finally, it constructs the detection
and survival covariate design matrices from the `capture_formula` and
`survival_formula` arguments, described in the [Including
detection/non-detecation covariates](#detection-covar) and [Including
survival covariates](#survival-covar) sections. The output provides data
frames as list elements, and a list element called `stan_d`, which
contains data that has been pre-processed for model fitting.

### Including detection/non-detection covariates {#detection-covar}

Survey-level covariates that might affect the probability of detection
can be included via the `capture_formula` argument to `clean_data`. Such
covariates must be in the `surveys` dataframe. Prior to using a column
containing continuous values as a covariate, it is important to rescale
these values to have mean = 0 and standard deviation = 1. Using a
covariate value like `year`, which typically has a mean in the thousands
and a standard deviation far from one may cause numerical issues when
fitting a model.

Covariate column names must be unique across all data frames supplied to
`clean_data()` (`captures`, `surveys`, `additions`, and `removals`).
Internally, `clean_data()` joins these data frames together using all of
their shared column names as the join key; if an added covariate shares
a name with a column in another data frame, it will unexpectedly become
part of that join and can silently produce incorrect results.

A detection covariate, such as `survey_duration`, can be included as
follows:

```{r detection-covariate, message=FALSE, eval=FALSE}
# center and scale the covariate
survey <- survey |>
  mutate(c_survey_duration = c(scale(survey_duration)))

# specify the formula
data <- clean_data(
  captures = capture, 
  surveys = survey,
  additions = translocation,
  capture_formula = ~ c_survey_duration
  )
```

### Including survival covariates {#survival-covar}

Individual-level survival covariates can be included via the
`survival_formula` and `survival_fill_value` arguments. Any variable
named in `survival_formula` must be a column in `captures` or
`additions` -- these are the only two places `clean_data()` looks for
individual-level values -- and must be fixed for each individual over
the entire study. The arguments must be specified together, because
`survival_fill_value` supplies the covariate value for every individual
the data says nothing about: the (typically numerous) augmented
pseudo-individuals, which were never captured and so can never have a
recorded value, and any real individual whose `captures`/`additions`
records lack the covariate. In other words, `survival_fill_value` should
be set to whichever level the covariate takes on *by default*, in the
absence of an explicit value.

For example, suppose you want to evaluate the effect of an experiment
that was conducted on a subset of individuals. In that experiment, a
treatment is applied to half of the individuals in the subset
("treated") and the other half are "untreated" controls. For individuals
that were not part of the experiment and for pseudo-individuals, a
`treatment` value is absent (`NA`). Because the pseudo-individuals stand
in for the general population that might exist but was never observed,
and because an absent value already means "untreated" for real
individuals that were not part of the experiment, the fill value should
almost always be "untreated":

```{r survival-covariate, message=FALSE, eval=FALSE}
# specify the formula
data <- clean_data(
  captures = capture, 
  surveys = survey,
  additions = translocation,
  survival_formula = ~ treatment,
  survival_fill_value = c(treatment = "untreated")
  )
```

Setting `survival_fill_value = c(treatment = "treated")` would only be
appropriate in the reverse situation, where "treated" is the default for
individuals that were not included in the experiment and for the control
individuals.

Pseudo-individuals make some survival covariates hard to include,
specifically when the covariate value assigned to a pseudo-individual
(via the `survival_fill_value`) is ambiguous. For example, adding a
`sex` covariate requires assigning a sex to every pseudo-individual, but
what sex to assign is ambiguous. One approach is to assign sex based on
sampling sex from the observed sex ratio, but this assumes that the
sexes have equal detectability. If this assumption is violated,
estimates will likely be biased. Given these challenges, the best
approach is include only covariates for which the `survival_fill_value`
is unambiguous (see example above).

## Model structure {#model-structure}

The `mrmr2` package implements a Bayesian open-population Jolly-Seber
mark-recapture model with known additions to and removals from the
population. The model tracks the states of $M$ individuals that comprise
a superpopulation made up of real and pseudo-individuals (see Joseph and
Knapp 2018, *Ecosphere* for details).

We assume a robust sampling design where the states of individuals are
constant within primary periods, and primary periods have repeat
secondary periods within them (during which observations are made). The
possible states of individuals include "not recruited", "alive", and
"dead". The possible observations of individuals include "detected" and
"not detected". We assume that individuals that are in the "not
recruited" or "dead" states are never detected (i.e., there are no
mistakes in the individual PIT tag ID records).

For reference, the parameters estimated by the model are as follows:

```{=html}
<style>
.param-table table {
  table-layout: fixed;
  width: 100%;
}
.param-table col:nth-child(2), .param-table col:nth-child(3) {
  width: 8%;
}
.param-table col:nth-child(4) {
  width: 55%;
}
</style>
```

::: param-table
| Parameter | Symbol | Model | Definition |
|------------------|------------------|------------------|------------------|
| `alpha_lambda` | $\alpha_\lambda$ | all | Intercept of per-primary-period recruitment probability, on the logit scale |
| `sigma_lambda` | $\sigma_\lambda$ | all | Standard deviation of the recruitment random effect (`eps_lambda`) across primary periods |
| `eps_lambda` | $\epsilon_\lambda$ | all | Standardized per-primary-period random effect on recruitment probability |
| `lambda` | $\lambda$ | all | Probability that an as-yet-unrecruited individual is recruited during a given primary period |
| `beta_phi` | $\beta_\phi$ | all | Coefficients (intercept and any survival covariates) for survival probability, on the logit scale |
| `sigma_phi` | $\sigma_\phi$ | all | Standard deviation of the survival random effect (`eps_phi`) across primary periods |
| `eps_phi` | $\epsilon_\phi$ | all | Per-primary-period random effect on survival probability |
| `phi` | $\phi$ | all | Probability that an alive individual survives from one primary period to the next |
| `beta_detect` | $\beta_p$ | all | Coefficients (intercept and any detection covariates) for detection probability, on the logit scale |
| `sigma_detect_period` | $\sigma_{p,\text{period}}$ | `detect_re` | Standard deviation of the between-primary-period detection random effect (`eps_detect_period`) |
| `eps_detect_period` | $\epsilon_{p,\text{period}}$ | `detect_re` | Per-primary-period random effect on detection probability, shared by all surveys in that period |
| `sigma_detect_occasion` | $\sigma_{p,\text{occasion}}$ | `detect_re` | Standard deviation of the within-primary-period (per-survey) detection random effect (`eps_detect_occasion`) |
| `eps_detect_occasion` | $\epsilon_{p,\text{occasion}}$ | `detect_re` | Per-survey random effect on detection probability, nested within primary period |
| `sigma_detect_indiv` | $\sigma_{p,\text{indiv}}$ | `indiv_re` | Standard deviation of the individual-level detection random effect (`eps_detect_indiv`) |
| `eps_detect_indiv` | $\epsilon_{p,\text{indiv}}$ | `indiv_re` | Per-individual random effect on detection probability (individual detection heterogeneity, $M_h$) |
| `s` | $s$ | all | Latent state of each individual (not yet recruited, alive, or dead) in each primary period |
| `N` | $N$ | all | Estimated population size (number of "alive" individuals) in each primary period |
| `B` | $B$ | all | Estimated number of newly recruited individuals in each primary period |
| `Nsuper` | $N_{super}$ | all | Estimated superpopulation size: the total number of individuals ever alive during the study |
:::

## Model fitting

To fit a mark-recapture model, use `mrmr2::fit_model()`. This model
accounts for known introductions into and removals from the population,
and has random effects to account for variation in survival and
recruitment through time. It is also possible to add random effects to
account for [variation in detection](#detection-re) at the period
(`primary period`), occasion (`secondary period`), and individual level.

The default number of chains is 4. Generally, at least 1000 warmup and
1000 sampling iterations should be used; the example model below uses
only 10 iterations of each to ensure a short run time.

```{r fit-model, message=FALSE, warning=FALSE, results='hide'}
# fit-model
mod <- fit_model(
  data = data, 
  chains = 4,
  iter_warmup = 10, 
  iter_sampling = 10
)
```

To reduce `mrmr2::fit_model()` run times, use within-chain
parallelization by providing the following additional arguments:

- `cpp_options = list(stan_threads = TRUE)` which compiles the model
  with threading support.
- `parallel_chains` specifies the number of chains to run in parallel,
  and must be $\le$ `chains`.
- `threads_per_chain` is set to an integer value that ensures that all
  physical cores are used. The number of `parallel_chains` ×
  `threads_per_chain` must equal the number of physical cores available.
  For example, on a machine with 8 physical cores, specify 4 chains, 4
  parallel chains, and 2 threads per chain. On a machine with 16
  physical cores, specify 4 chains, 4 parallel chains, and 4 threads per
  chain. In the following code chunk, threads per chain is set
  automatically based on the number of cores available.

```{r fit-model-threaded, message=FALSE, warning=FALSE, eval=FALSE}
# fit-model-threaded
# WARNING: This model run takes 10-20 minutes
mod <- fit_model(
  data = data, 
  chains = 4,
  parallel_chains = 4,
  threads_per_chain = max(1, parallel::detectCores() %/% 4), 
  cpp_options = list(stan_threads = TRUE),
  iter_warmup = 1000, 
  iter_sampling = 1000
)
```

After the `mrmr2::fit_model` step completes, diagnose potential
non-convergence of the MCMC algorithm by inspecting traceplots and Rhat
values. Note that `mod` from the `fit-model` code chunk above is fit
with only 10 warmup and 10 sampling iterations, to minimize the run time
of this step. As a result of the limited number of iterations, the
resulting traceplots and Rhat values are not informative. To provide
more useful diagnostics, the traceplots and Rhat table below are based
on `mod` from the `fit_model_threaded` code chunk immediately above,
which is a more typical `fit_model` specification. To avoid this code
chunk's 10-20 minute run time, it was pre-run and the saved model fit
was used to create the traceplots and Rhat table.

```{r regenerate-traceplot-cache, include=FALSE, message=FALSE, warning=FALSE}
# Maintainer-only setup chunk, hidden from the rendered vignette. It
# creates `draws_for_plot`, used by the plot-traceplots-prerun and
# print-model-prerun chunks below, from a cached RDS file.
#
# To refresh that cache (e.g., after the example data or the Stan model
# change), delete inst/extdata/traceplot-draws-example.rds and re-knit
# this vignette from the package source tree (e.g., RStudio Knit button). 
# This chunk will then re-fit the model with the same settings shown in the
# fit-model-threaded chunk above (~20 minutes) and re-save the cache;
# if the file is already present, re-fitting is skipped and the cache
# just loads (fast, as in a normal build).
pars_to_plot <- c('alpha_lambda',
                  'sigma_lambda',
                  'beta_phi',
                  'sigma_phi',
                  'beta_detect')

cache_path <- "../inst/extdata/traceplot-draws-example.rds"

if (!file.exists(cache_path)) {
  cache_mod <- fit_model(
    data = data,
    chains = 4,
    parallel_chains = 4,
    threads_per_chain = max(1, parallel::detectCores() %/% 4),
    cpp_options = list(stan_threads = TRUE),
    iter_warmup = 1000,
    iter_sampling = 1000
  )
  saveRDS(cache_mod$m_fit$draws(pars_to_plot), cache_path)
}

draws_for_plot <- readRDS(cache_path)
```

```{r plot-traceplots-prerun, echo=FALSE}
bayesplot::mcmc_trace(draws_for_plot)
```

Signs of potential lack of convergence include any moments (e.g., mean
or standard deviation) changing over time, and high within-chain
auto-correlation. High within-chain autocorrelation means the sampler is
exploring the posterior slowly (poor mixing), which reduces the
effective sample size relative to the raw number of draws, and shows up
in a trace plot as slow-moving, snake-like trends rather than a "hairy
caterpillar" pattern that jumps around its mean quickly. Additionally,
if multiple chains initialized with different starting values do not
converge to similar equilibrium values, this may also suggest issues
with convergence. These traceplots show typical "hairy caterpillar"
patterns indicative of model convergence.

Here is the Rhat table. Rhat values $\ge$ 1.01 indicate of a lack of
convergence.

```{r print-model-prerun, echo=FALSE}
posterior::summarise_draws(draws_for_plot)
```

If you run the `fit_model_threaded` chunk yourself, you can generate the
same diagnostics for your fit as follows:

```{r plot-traceplots, eval=FALSE}
pars_to_plot <- c('alpha_lambda',
                  'sigma_lambda',
                  'beta_phi',
                  'sigma_phi',
                  'beta_detect')

bayesplot::mcmc_trace(mod$m_fit$draws(pars_to_plot))
```

```{r print-model, eval=FALSE}
mod$m_fit$summary(pars_to_plot)
```

## Detection random-effect models {#detection-re}

By default (`model = "base"`), detection probability is a purely
deterministic function of any covariates supplied via `capture_formula`
to `clean_data()`. With the default `capture_formula = ~ 1`, every
survey in the dataset shares one scalar detection probability. Survival
and recruitment, by contrast, already have a random effect that lets
them vary by primary period. `mrmr2::fit_model()` offers two optional
extensions, selected with the `model` argument, that add an analogous
random effect to detection:

- `model = "detect_re"` adds a between-primary-period random effect and
  a within-primary-period (per-survey) random effect to detection
  probability. Simulation work found this mainly corrects overconfident
  (too-narrow) uncertainty intervals on abundance (`N`) and recruitment
  (`B`) when true detectability drifts across periods or survey
  occasions — it has little effect on the point estimates themselves.
- `model = "indiv_re"` adds an individual-level random effect to account
  for some individuals being persistently harder to detect than others
  (e.g., due to body size, behavior, or microhabitat use). Unlike
  `"detect_re"`, ignoring real individual heterogeneity of this kind
  causes a genuine *negative bias* in point estimates of `N` —
  chronically hard-to-detect individuals get undercounted — matching
  classical closed-population heterogeneity (`M_h`) theory.

The following code chunk adds an individual-level random effect to
detection probability

```{r fit-model-re, eval=FALSE}
mod_indiv_re <- fit_model(
  data = data,
  model = "indiv_re",
  chains = 4,
  parallel_chains = 4,
  threads_per_chain = max(1, parallel::detectCores() %/% 4), 
  cpp_options = list(stan_threads = TRUE),
  iter_warmup = 1000,
  iter_sampling = 1000
)
```

These extensions are not free. Both take longer to fit than `"base"`,
and `"indiv_re"` in particular produces substantially larger saved fits,
since its random effect has one entry per individual in the augmented
superpopulation (`M`, typically several times the number of animals ever
observed) rather than one per primary period or survey. `"detect_re"`'s
two variance components are only weakly separable, which can trigger
divergent transitions when one of them is at or near zero; `fit_model()`
handles this automatically by defaulting `adapt_delta` to `0.95` (rather
than cmdstanr's usual `0.8`) whenever `model = "detect_re"` and
`adapt_delta` is left unspecified.

One more caveat when interpreting a `"detect_re"` fit: with only a few
secondary surveys per primary period, the model can attribute some of
what is really per-survey (`sigma_detect_occasion`) noise to the
between-period term (`sigma_detect_period`) instead, purely because a
small sample of occasion-level draws can look like a period-level shift
by chance. Before concluding there is genuine period-to-period variation
in detectability, compare the posterior for `sigma_detect_period`
against the noise floor it would show even with no real period effect,
roughly `sigma_detect_occasion / sqrt(J)`, where `J` is the number of
secondary surveys in a primary period. This between-period versus
survey-level ambiguity doesn't bias `N`, `B`, `phi`, or `lambda` — those
depend only on the sum of the two random effects — but it does mean the
two `sigma` estimates shouldn't be read in isolation.

Additional details about the `detect_re` and `indiv_re` models,
including evaluation of outputs from model runs using simulated and real
data, are provided in [An Introduction to CMR Analysis Using the mrmr
and mrmr2
Packages](https://github.com/SNARL1/mrmr-revisited/blob/main/introduction-to-cmr-with-mrmr.qmd).

## Built-in visualizations

Time series of population abundance, recruitment, and survival of
introduced cohorts are available through the `mrmr2::plot_model()`
function.

```{r plot-abundance}
plot_model(mod, what = 'abundance')
```

```{r plot-recruitment}
plot_model(mod, what = 'recruitment')
```

```{r plot-survival}
plot_model(mod, what = 'survival')
```

`what = 'survival'` only pulls posterior draws of the latent state (`s`)
for translocated individuals, rather than for every individual (real and
data-augmented pseudo-individual) tracked by the model (behavior of
`what = 'survival'` in `mrmr`). On sites with a large superpopulation
relative to the number of translocated individuals, this keeps memory
use for this plot far below what it would be if the full `s` array were
pulled and then filtered down afterward (see [An Introduction to CMR
Analysis Using the mrmr and mrmr2
Packages](https://github.com/SNARL1/mrmr-revisited/blob/main/introduction-to-cmr-with-mrmr.qmd)
for details).

## Custom visualizations

The default plots can be customized using standard plotting functions:

```{r plots-custom, fig.height=7, fig.width=7}
p1 <- plot_model(mod, what = "abundance") 
p1 <- p1 + ggtitle("A. Population size") + 
  ylab("Number of adults") + 
  expand_limits(y = 0) +
  theme_classic()
  
p2 <- plot_model(mod, what = "recruitment")
p2 <- p2 + ggtitle("B. Recruitment") +
  ylab("Number of adults") +
  theme_classic()

p3 <- plot_model(mod, what = "survival")
p3 <- p3 + ggtitle("C. Survival of Translocated Cohorts") +
  ylab("Number of adults") + 
  theme_classic()
  
p1 / p2 /p3
```

To save the custom plots (in .png, .pdf, and other formats):

```{r save-plot, eval=FALSE}
ggsave("custom_plots.png", height = 6.5, width = 6.5)
```

### Plotting other parameters

Any of the plotting functionality that you would expect from a `stanfit`
model is available, by accessing the `m_fit` list element from a model
object. For example, we could assess the posterior for the
superpopulation size:

```{r plot-nsuper, eval=FALSE}
bayesplot::mcmc_trace(mod$m_fit$draws("Nsuper"))
```

See [Model structure](#model-structure) for parameter names and
definitions.

## Translocation/reintroduction survival tables

If individuals were added to sites, you can generate a table of
estimated survival through time for each introduced cohort (release
date) as a function of years since introduction:

```{r survival-table}
survival_table(mod, by_cohort = TRUE)
```

You can also get estimates of survival through time by individual:

```{r survival-table-ind}
survival_table(mod, by_individual = TRUE)
```
